---
title: "RNA-seq Project - Initial Setup and Verification"
author: "Jose Hidalgo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)

# Define project root directory
project_dir <- "C:/Users/10935/Documents/Proyecto_Adenocarcinoma_RNAseq"
knitr::opts_knit$set(root.dir = project_dir)
```

# Introduction

This document verifies the initial structure of the RNA-seq analysis project for biomarker detection in lung adenocarcinoma (LUAD).

**Experimental design:**
- 10 tumor samples (NSCLC tissue)
- 10 normal samples (non-malignant tissue)
- Paired-end, unstranded sequencing
- Platform: Illumina HiSeq 2500

---

# 1. Confirm Project Directory
```{r confirm-directory}
cat("Project directory:", getwd(), "\n")
cat("Directory configured correctly\n")
```

---

# 2. Verify FASTQ Files
```{r verify-fastq}
fastq_dir <- "00_raw_data/fastq"
fastq_files <- list.files(fastq_dir, pattern = "\\.fastq\\.gz$", full.names = FALSE)

cat("FASTQ FILES\n")

if (length(fastq_files) > 0) {
  # Count R1 and R2 files (paired-end)
  r1_files <- grep("_1\\.fastq\\.gz$", fastq_files, value = TRUE)
  r2_files <- grep("_2\\.fastq\\.gz$", fastq_files, value = TRUE)

  cat("Total FASTQ files:", length(fastq_files), "\n")
  cat("  R1 files (forward):", length(r1_files), "\n")
  cat("  R2 files (reverse):", length(r2_files), "\n")
  cat("  Total samples:", length(r1_files), "\n\n")

} else {
  cat("No FASTQ files found\n")
}
```

---

# 2.1. Decompress FASTQ Files (Optional)

**Note:** Most bioinformatics tools read `.fastq.gz` files directly without decompression. Only decompress if required by a specific tool.

**To run:** change `eval=FALSE` to `eval=TRUE`.
```{r decompress-fastq, eval=FALSE}
library(R.utils)

fastq_dir <- "00_raw_data/fastq"
fastq_gz_files <- list.files(fastq_dir, pattern = "\\.fastq\\.gz$", full.names = TRUE)

cat("Decompressing", length(fastq_gz_files), "FASTQ files...\n")
cat("This may take 30-60 minutes\n\n")

for (i in seq_along(fastq_gz_files)) {
  file_gz <- fastq_gz_files[i]
  cat(sprintf("[%d/%d] %s\n", i, length(fastq_gz_files), basename(file_gz)))
  R.utils::gunzip(file_gz, remove = FALSE, overwrite = TRUE)
}

cat("\nDecompression complete\n")
```

---

# 3. Metadata Preparation for DESeq2

## 3.1. Read Source Files
```{r read-metadata}
library(tidyverse)

metadata_dir <- "00_raw_data/metadata"

# Read CSV files
sra_table   <- read.csv(file.path(metadata_dir, "SraRunTable.csv"),
                        stringsAsFactors = FALSE)
samplesheet <- read.csv(file.path(metadata_dir, "samplesheet.csv"),
                        stringsAsFactors = FALSE)

cat("SraRunTable.csv:", nrow(sra_table), "samples\n")
cat("samplesheet.csv:", nrow(samplesheet), "samples\n")
```

## 3.2. Build Consolidated Metadata
```{r create-consolidated-metadata}
# Build consolidated metadata for DESeq2
metadata <- sra_table %>%
  select(Run, source_name, Histology, gender, AGE, stage_tnm) %>%
  mutate(
    # Condition column: Normal as reference level for DESeq2
    condition = ifelse(grepl("NSCLC", source_name), "Tumor", "Normal"),
    condition = factor(condition, levels = c("Normal", "Tumor")),

    # Sample ID
    sample_id = Run,

    # Histology grade as factor
    histology_grade = ifelse(is.na(Histology), "NA", as.character(Histology)),
    histology_grade = factor(histology_grade)
  ) %>%
  # Join FASTQ file paths from samplesheet
  left_join(samplesheet %>% select(sample, fastq_1, fastq_2, strandedness),
            by = c("Run" = "sample")) %>%
  # Fix FASTQ path prefixes
  mutate(
    fastq_1 = gsub("raw_data_fastq/", "00_raw_data/fastq/", fastq_1),
    fastq_2 = gsub("raw_data_fastq/", "00_raw_data/fastq/", fastq_2)
  ) %>%
  # Reorder columns
  select(sample_id, condition, fastq_1, fastq_2, strandedness,
         gender, AGE, histology_grade, stage_tnm, source_name)

# Summary
cat("CONSOLIDATED METADATA FOR DESeq2\n")
cat("Total samples:", nrow(metadata), "\n")
cat("\nSamples per condition:\n")
print(table(metadata$condition))

cat("\nFirst 3 tumor samples:\n")
print(head(metadata %>% filter(condition == "Tumor") %>%
             select(sample_id, condition, gender, AGE, histology_grade), 3))

cat("\nFirst 3 normal samples:\n")
print(head(metadata %>% filter(condition == "Normal") %>%
             select(sample_id, condition, fastq_1), 3))

# Save consolidated metadata
write.csv(metadata,
          "00_raw_data/metadata/metadata_consolidado.csv",
          row.names = FALSE, quote = FALSE)

# Minimal version for DESeq2
metadata_deseq <- metadata %>%
  select(sample_id, condition, fastq_1, fastq_2)

write.csv(metadata_deseq,
          "00_raw_data/metadata/metadata_deseq2.csv",
          row.names = FALSE, quote = FALSE)

cat("\nFiles saved:\n")
cat("  - metadata_consolidado.csv (full metadata)\n")
cat("  - metadata_deseq2.csv (minimal, for DESeq2)\n")
```

---

# 4. Verify Reference Genome and Annotation
```{r verify-reference}
# Genome FASTA
genome_file <- "03_genome_reference/genome_fasta/GRCh38.primary_assembly.genome.fa.gz"
if (file.exists(genome_file)) {
  cat("Genome found (compressed)\n")
  cat("  Size:", round(file.info(genome_file)$size / (1024^2), 2), "MB\n")
} else {
  cat("Genome NOT found\n")
  cat("  Download from:\n")
  cat("  https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_47/GRCh38.primary_assembly.genome.fa.gz\n")
}

# GTF annotation
cat("\n")
gtf_file <- "03_genome_reference/annotation/gencode.v47.primary_assembly.annotation.gtf.gz"
if (file.exists(gtf_file)) {
  cat("GTF annotation found (compressed)\n")
  cat("  Size:", round(file.info(gtf_file)$size / (1024^2), 2), "MB\n")
} else {
  cat("GTF NOT found\n")
  cat("  Download from:\n")
  cat("  https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_47/gencode.v47.primary_assembly.annotation.gtf.gz\n")
}
```

---

# 5. Decompress Reference Files

**To run:** change `eval=FALSE` to `eval=TRUE`.
```{r decompress-reference, eval=FALSE}
if (!require("R.utils")) install.packages("R.utils")
library(R.utils)

cat("Decompressing reference files...\n\n")

# Decompress genome FASTA
genome_gz <- "03_genome_reference/genome_fasta/GRCh38.primary_assembly.genome.fa.gz"
if (file.exists(genome_gz)) {
  cat("Decompressing genome (may take several minutes)...\n")
  R.utils::gunzip(genome_gz, remove = FALSE, overwrite = TRUE)
  cat("Genome decompressed\n\n")
}

# Decompress GTF
gtf_gz <- "03_genome_reference/annotation/gencode.v47.primary_assembly.annotation.gtf.gz"
if (file.exists(gtf_gz)) {
  cat("Decompressing GTF...\n")
  R.utils::gunzip(gtf_gz, remove = FALSE, overwrite = TRUE)
  cat("GTF decompressed\n\n")
}

cat("Decompression complete\n")
```

---

# 6. Final Summary
```{r summary}
# Check all components
fastq_ok    <- length(fastq_files) >= 40
metadata_ok <- file.exists("00_raw_data/metadata/metadata_consolidado.csv")
genome_ok   <- file.exists(genome_file)
gtf_ok      <- file.exists(gtf_file)

checks <- data.frame(
  Component = c("FASTQ files", "Consolidated metadata",
                "Reference genome", "GTF annotation"),
  Status    = c(
    ifelse(fastq_ok,    "OK", "MISSING"),
    ifelse(metadata_ok, "OK", "MISSING"),
    ifelse(genome_ok,   "OK", "MISSING"),
    ifelse(gtf_ok,      "OK", "MISSING")
  ),
  stringsAsFactors = FALSE
)

print(checks)

cat("\n")
if (all(checks$Status == "OK")) {
  cat("INITIAL VERIFICATION COMPLETE\n\n")
  cat("Next steps:\n")
  cat("  1. Decompress genome and GTF (Section 5)\n")
  cat("  2. Quality control with FastQC (Script 01)\n")
  cat("  3. Build index and align with Rsubread (Script 02)\n")
} else {
  cat("Complete missing components before proceeding.\n")
}
```

---

# Session Info
```{r session-info}
sessionInfo()
```
