---
title: "Functional Analysis: GO, KEGG, and GSEA"
author: "Jose Hidalgo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE,
                      fig.width = 10, fig.height = 8)
project_dir <- "C:/Users/10935/Documents/Proyecto_Adenocarcinoma_RNAseq"
knitr::opts_knit$set(root.dir = project_dir)
```

# 1. Load Libraries
```{r load-libraries, message=FALSE, warning=FALSE}
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(tidyverse)
library(ggplot2)

cat("clusterProfiler version:", as.character(packageVersion("clusterProfiler")), "\n")
cat("org.Hs.eg.db version:   ", as.character(packageVersion("org.Hs.eg.db")), "\n")
```

# 2. Load DEGs
```{r load-degs}
degs        <- read.csv("06_differential_expression/DEGs_significant.csv")
all_results <- read.csv("06_differential_expression/DESeq2_all_results.csv")

cat("Total DEGs:    ", nrow(degs), "\n")
cat("Upregulated:   ", sum(degs$log2FoldChange > 0), "\n")
cat("Downregulated: ", sum(degs$log2FoldChange < 0), "\n")
```

# 3. ID Conversion: Ensembl to ENTREZ
```{r convert-ids}
# Strip version suffixes from Ensembl IDs (e.g. ENSG00000001.12 -> ENSG00000001)
degs$gene_id_clean        <- sub("\\..*", "", degs$gene_id)
all_results$gene_id_clean <- sub("\\..*", "", all_results$gene_id)

# Map Ensembl to ENTREZ for DEGs
degs$entrez <- mapIds(
  org.Hs.eg.db,
  keys     = degs$gene_id_clean,
  column   = "ENTREZID",
  keytype  = "ENSEMBL",
  multiVals = "first"
)

# Map Ensembl to ENTREZ for all results (needed for GSEA ranked list)
all_results$entrez <- mapIds(
  org.Hs.eg.db,
  keys     = all_results$gene_id_clean,
  column   = "ENTREZID",
  keytype  = "ENSEMBL",
  multiVals = "first"
)

# Keep only DEGs with a valid ENTREZ ID
degs_with_entrez <- degs %>% filter(!is.na(entrez))

cat("\nDEGs with ENTREZ ID:    ", nrow(degs_with_entrez), "\n")
cat("DEGs without ENTREZ ID: ", sum(is.na(degs$entrez)), "\n")
```

# 4. Gene Ontology — Biological Process

**To run:** change `eval=FALSE` to `eval=TRUE`.
```{r go-bp, eval=FALSE}
cat("Running GO Biological Process enrichment...\n")

ego_bp <- enrichGO(
  gene          = degs_with_entrez$entrez,
  OrgDb         = org.Hs.eg.db,
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2,
  readable      = TRUE   # convert ENTREZ IDs to gene symbols in output
)

saveRDS(ego_bp, "07_functional_analysis/GO_BP_results.rds")
write.csv(as.data.frame(ego_bp), "07_functional_analysis/GO_BP_results.csv", row.names = FALSE)

cat("GO-BP terms enriched:", nrow(ego_bp), "\n")
```

# 5. Gene Ontology — Molecular Function
```{r go-mf, eval=FALSE}
cat("Running GO Molecular Function enrichment...\n")

ego_mf <- enrichGO(
  gene          = degs_with_entrez$entrez,
  OrgDb         = org.Hs.eg.db,
  ont           = "MF",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2,
  readable      = TRUE
)

saveRDS(ego_mf, "07_functional_analysis/GO_MF_results.rds")
write.csv(as.data.frame(ego_mf), "07_functional_analysis/GO_MF_results.csv", row.names = FALSE)

cat("GO-MF terms enriched:", nrow(ego_mf), "\n")
```

# 6. Gene Ontology — Cellular Component
```{r go-cc, eval=FALSE}
cat("Running GO Cellular Component enrichment...\n")

ego_cc <- enrichGO(
  gene          = degs_with_entrez$entrez,
  OrgDb         = org.Hs.eg.db,
  ont           = "CC",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2,
  readable      = TRUE
)

saveRDS(ego_cc, "07_functional_analysis/GO_CC_results.rds")
write.csv(as.data.frame(ego_cc), "07_functional_analysis/GO_CC_results.csv", row.names = FALSE)

cat("GO-CC terms enriched:", nrow(ego_cc), "\n")
```

# 7. KEGG Pathway Enrichment
```{r kegg, eval=FALSE}
cat("Running KEGG pathway enrichment...\n")

# Increase timeout — KEGG requires internet access
options(download.file.method = "libcurl")
options(timeout = 300)

kegg <- enrichKEGG(
  gene          = degs_with_entrez$entrez,
  organism      = "hsa",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

# Replace ENTREZ IDs with gene symbols in output
kegg <- setReadable(kegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

saveRDS(kegg, "07_functional_analysis/KEGG_results.rds")
write.csv(as.data.frame(kegg), "07_functional_analysis/KEGG_results.csv", row.names = FALSE)

cat("KEGG pathways enriched:", nrow(kegg), "\n")
```

# 8. Load Results
```{r load-results}
if (file.exists("07_functional_analysis/GO_BP_results.rds")) {
  ego_bp <- readRDS("07_functional_analysis/GO_BP_results.rds")
  cat("GO-BP terms:    ", nrow(ego_bp), "\n")
}
if (file.exists("07_functional_analysis/GO_MF_results.rds")) {
  ego_mf <- readRDS("07_functional_analysis/GO_MF_results.rds")
  cat("GO-MF terms:    ", nrow(ego_mf), "\n")
}
if (file.exists("07_functional_analysis/GO_CC_results.rds")) {
  ego_cc <- readRDS("07_functional_analysis/GO_CC_results.rds")
  cat("GO-CC terms:    ", nrow(ego_cc), "\n")
}
if (file.exists("07_functional_analysis/KEGG_results.rds")) {
  kegg <- readRDS("07_functional_analysis/KEGG_results.rds")
  cat("KEGG pathways:  ", nrow(kegg), "\n")
}
```

# 9. Visualization: GO-BP
```{r plot-go-bp, fig.width=14, fig.height=10}
if (exists("ego_bp") && nrow(ego_bp) > 0) {
  p1 <- dotplot(ego_bp, showCategory = 20) +
    ggtitle("GO Biological Process: Top 20") +
    scale_y_discrete(labels = function(x) str_wrap(x, width = 40)) +
    theme(axis.text.y = element_text(size = 10),
          plot.title  = element_text(size = 13, face = "bold"))
  print(p1)
  ggsave("08_figures_tables/article_figures/Fig3A_GO_BP.png",
         p1, width = 14, height = 10, dpi = 300)
}
```

# 10. Visualization: KEGG
```{r plot-kegg, fig.width=12, fig.height=8}
if (exists("kegg") && nrow(kegg) > 0) {
  p3 <- dotplot(kegg, showCategory = 20) +
    ggtitle("KEGG Pathways: Top 20")
  print(p3)
  ggsave("08_figures_tables/KEGG_dotplot.png", p3, width = 12, height = 8, dpi = 300)

  p4 <- barplot(kegg, showCategory = 15) +
    ggtitle("KEGG Pathways: Top 15")
  print(p4)
  ggsave("08_figures_tables/KEGG_barplot.png", p4, width = 10, height = 8, dpi = 300)
}
```

# 11. GSEA — Gene Set Enrichment Analysis

**To run:** change `eval=FALSE` to `eval=TRUE`.
```{r gsea, eval=FALSE}
cat("Running GSEA...\n")

# Build ranked gene list using all genes (not just DEGs)
# Rank metric: log2FoldChange — positive = upregulated in tumor
all_genes <- all_results %>%
  filter(!is.na(entrez), !is.na(log2FoldChange)) %>%
  # Remove duplicate ENTREZ IDs: keep the one with highest |log2FC|
  group_by(entrez) %>%
  slice_max(abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(desc(log2FoldChange))

gene_list        <- all_genes$log2FoldChange
names(gene_list) <- all_genes$entrez

cat("Genes in ranked list: ", length(gene_list), "\n")
cat("Unique ENTREZ IDs:    ", length(unique(names(gene_list))), "\n")

# GSEA on GO Biological Process terms
gsea_go <- gseGO(
  geneList      = gene_list,
  OrgDb         = org.Hs.eg.db,
  ont           = "BP",
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE
)

saveRDS(gsea_go, "07_functional_analysis/GSEA_GO_results.rds")
write.csv(as.data.frame(gsea_go), "07_functional_analysis/GSEA_GO_results.csv", row.names = FALSE)

# GSEA on KEGG pathways
gsea_kegg <- gseKEGG(
  geneList      = gene_list,
  organism      = "hsa",
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE
)

gsea_kegg <- setReadable(gsea_kegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

saveRDS(gsea_kegg, "07_functional_analysis/GSEA_KEGG_results.rds")
write.csv(as.data.frame(gsea_kegg), "07_functional_analysis/GSEA_KEGG_results.csv", row.names = FALSE)

cat("GSEA GO terms:      ", nrow(gsea_go), "\n")
cat("GSEA KEGG pathways: ", nrow(gsea_kegg), "\n")
```

# 12. Visualization: GSEA
```{r plot-gsea, fig.width=16, fig.height=12}
if (file.exists("07_functional_analysis/GSEA_GO_results.rds")) {
  gsea_go <- readRDS("07_functional_analysis/GSEA_GO_results.rds")

  if (nrow(gsea_go) > 0) {
    p5 <- dotplot(gsea_go, showCategory = 15, split = ".sign") +
      facet_grid(~.sign) +
      ggtitle("GSEA: GO Biological Process") +
      theme(axis.text.y  = element_text(size = 9),
            strip.text   = element_text(size = 11, face = "bold"),
            plot.title   = element_text(size = 13, face = "bold")) +
      scale_y_discrete(labels = function(x) str_wrap(x, width = 35))
    print(p5)
    ggsave("08_figures_tables/article_figures/Fig3C_GSEA_GO.png",
           p5, width = 16, height = 12, dpi = 300)
  }
}

if (file.exists("07_functional_analysis/GSEA_KEGG_results.rds")) {
  gsea_kegg <- readRDS("07_functional_analysis/GSEA_KEGG_results.rds")

  if (nrow(gsea_kegg) > 0) {
    p6 <- dotplot(gsea_kegg, showCategory = 20, split = ".sign") +
      facet_grid(~.sign) +
      ggtitle("GSEA: KEGG Pathways")
    print(p6)
    ggsave("08_figures_tables/GSEA_KEGG_dotplot.png", p6, width = 14, height = 8, dpi = 300)
  }
}
```

# 13. Summary
```{r summary}
cat("FUNCTIONAL ANALYSIS SUMMARY\n\n")
cat("DEGs analyzed: ", nrow(degs_with_entrez), "\n\n")

if (exists("ego_bp")) cat("GO-BP terms:    ", nrow(ego_bp), "\n")
if (exists("ego_mf")) cat("GO-MF terms:    ", nrow(ego_mf), "\n")
if (exists("ego_cc")) cat("GO-CC terms:    ", nrow(ego_cc), "\n")
if (exists("kegg"))   cat("KEGG pathways:  ", nrow(kegg), "\n")

if (file.exists("07_functional_analysis/GSEA_GO_results.rds")) {
  gsea_go <- readRDS("07_functional_analysis/GSEA_GO_results.rds")
  cat("GSEA GO terms:  ", nrow(gsea_go), "\n")
}
if (file.exists("07_functional_analysis/GSEA_KEGG_results.rds")) {
  gsea_kegg <- readRDS("07_functional_analysis/GSEA_KEGG_results.rds")
  cat("GSEA KEGG:      ", nrow(gsea_kegg), "\n")
}

cat("\nOutput files:\n")
cat("  - GO_BP_results.csv\n")
cat("  - GO_MF_results.csv\n")
cat("  - GO_CC_results.csv\n")
cat("  - KEGG_results.csv\n")
cat("  - GSEA_GO_results.csv\n")
cat("  - GSEA_KEGG_results.csv\n")
cat("  + Figures in 08_figures_tables/\n")

cat("\nRNA-seq pipeline complete.\n")
```

```{r session-info}
sessionInfo()
```
