---
title: "WGCNA Analysis - LUAD"
author: "Jose Hidalgo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
project_dir <- "C:/Users/10935/Documents/Proyecto_Adenocarcinoma_RNAseq"
knitr::opts_knit$set(root.dir = project_dir)
```

# 1. Load Libraries
```{r libraries}
library(WGCNA)
library(DESeq2)
library(tidyverse)

# Enable multi-threading for WGCNA correlation computations
allowWGCNAThreads()
```

# 2. Load Data
```{r load-data}
# Load DESeq2 object saved in Script 04
dds <- readRDS("06_differential_expression/deseq2_object.rds")

# Apply variance stabilizing transformation for WGCNA
# blind = FALSE: uses design information to avoid over-shrinkage
vsd <- vst(dds, blind = FALSE)
expr_matrix <- assay(vsd)

# WGCNA expects samples as rows and genes as columns
datExpr <- t(expr_matrix)

dim(datExpr)
```

# 3. Sample Quality Control
```{r sample-qc}
# Check for genes or samples with excessive missing values
# allOK = TRUE means no filtering needed
gsg <- goodSamplesGenes(datExpr, verbose = 3)

gsg$allOK
```

# 4. Sample Clustering
```{r sample-clustering, fig.width=12, fig.height=6}
# Hierarchical clustering to detect outlier samples before network construction
sampleTree <- hclust(dist(datExpr), method = "average")

plot(sampleTree,
     main    = "Sample Clustering to Detect Outliers",
     sub     = "",
     xlab    = "",
     cex.lab = 1.2,
     cex.axis = 1,
     cex.main = 1.4)
```

# Outlier removal rationale
# SRR3475130 and SRR3475138 were identified as outliers in both PCA (separation
# along PC2) and hierarchical clustering (branch height ~160 vs ~100 for other samples).
# They were removed before WGCNA because co-expression network construction via Pearson
# correlation is more sensitive to outliers than DESeq2's negative binomial model.
# Both samples had been flagged in the differential expression analysis as transcriptomically
# atypical, consistent with the intratumoral heterogeneity characteristic of LUAD.

# 5. Remove Outliers
```{r remove-outliers}
outliers <- c("SRR3475130", "SRR3475138")
datExpr  <- datExpr[!rownames(datExpr) %in% outliers, ]

# Confirm removal
dim(datExpr)
rownames(datExpr)
```

# 6. Soft Thresholding Power Selection
```{r soft-threshold, fig.width=12, fig.height=6}
# Test powers 1-30 to find the one that achieves scale-free topology
powers <- c(1:30)

sft <- pickSoftThreshold(datExpr,
                         powerVector = powers,
                         verbose     = 5,
                         networkType = "signed")

# Plot scale-free fit and mean connectivity vs power
par(mfrow = c(1, 2))

# R² vs power — select the lowest power above the 0.80 threshold
plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit (R²)",
     main = "Scale Independence",
     type = "n")
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, col = "red")
abline(h = 0.80, col = "blue")

# Mean connectivity vs power — should decrease monotonically
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     main = "Mean Connectivity",
     type = "n")
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers, col = "red")
```

# 6.5 Filter to DEGs Only
```{r filter-degs}
# Restrict WGCNA to DEGs only — reduces noise and computation time
degs <- read.csv("06_differential_expression/DEGs_significant.csv")

# Clean version suffixes from Ensembl IDs
degs$gene_id_clean  <- sub("\\..*", "", degs$gene_id)
gene_ids_clean      <- sub("\\..*", "", colnames(datExpr))

# Keep only columns (genes) that are significant DEGs
deg_indices        <- which(gene_ids_clean %in% degs$gene_id_clean)
datExpr_filtered   <- datExpr[, deg_indices]

dim(datExpr_filtered)
cat("Genes in analysis:", ncol(datExpr_filtered), "\n")
```

# 7. Network Construction and Module Detection
```{r network, cache=TRUE}
# Selected power: 13 (first value achieving R² >= 0.80)
softPower <- 13

# Override R's built-in cor() with WGCNA's version to avoid conflicts
cor <- WGCNA::cor

net <- blockwiseModules(
  datExpr_filtered,
  power            = softPower,
  networkType      = "signed",    # signed: preserves direction of correlation
  TOMType          = "signed",
  minModuleSize    = 30,          # minimum genes per module
  mergeCutHeight   = 0.25,        # threshold for merging similar modules
  deepSplit        = 2,
  numericLabels    = FALSE,       # use color names for modules
  pamRespectsDendro = FALSE,
  saveTOMs         = FALSE,
  verbose          = 3
)

# Module sizes
table(net$colors)
```

# 8. Module-Trait Correlation
```{r module-trait, fig.width=10, fig.height=8}
# Encode condition as binary trait: 0 = Normal, 1 = Tumor
samples   <- rownames(datExpr_filtered)
condition <- ifelse(grepl("SRR3474", samples), "Normal", "Tumor")
traitData <- data.frame(
  Tumor     = as.numeric(condition == "Tumor"),
  row.names = samples
)

# Calculate module eigengenes (first PC of each module's expression)
MEs <- moduleEigengenes(datExpr_filtered, net$colors)$eigengenes
MEs <- orderMEs(MEs)

# Correlate eigengenes with Tumor trait
moduleTraitCor    <- cor(MEs, traitData, use = "p")
moduleTraitPvalue <- corPvalueStudent(moduleTraitCor, nrow(datExpr_filtered))

# Build annotation matrix for heatmap
textMatrix     <- paste0(signif(moduleTraitCor, 2), "\n(", signif(moduleTraitPvalue, 1), ")")
dim(textMatrix) <- dim(moduleTraitCor)

par(mar = c(6, 8.5, 3, 3))
labeledHeatmap(
  Matrix        = moduleTraitCor,
  xLabels       = "Tumor",
  yLabels       = names(MEs),
  ySymbols      = names(MEs),
  colorLabels   = FALSE,
  colors        = blueWhiteRed(50),
  textMatrix    = textMatrix,
  setStdMargins = FALSE,
  cex.text      = 0.8,
  zlim          = c(-1, 1),
  main          = "Module-Trait Relationships"
)
```

# 9. Extract Tumor-Associated Module Genes
```{r tumor-modules}
# Selected modules based on module-trait correlation:
# blue (activated), turquoise (suppressed), brown (activated)
tumor_modules <- c("blue", "turquoise", "brown")

module_genes <- data.frame(
  gene_id = colnames(datExpr_filtered),
  module  = net$colors
) %>%
  filter(module %in% tumor_modules)

table(module_genes$module)

write.csv(module_genes,
          "07_functional_analysis/WGCNA_tumor_module_genes.csv",
          row.names = FALSE)

cat("Total genes in tumor modules:", nrow(module_genes), "\n")
```

# 10. DEG and Module Overlap
```{r deg-module-overlap}
# Clean IDs for join
module_genes$gene_id_clean <- sub("\\..*", "", module_genes$gene_id)
degs$gene_id_clean         <- sub("\\..*", "", degs$gene_id)

# Inner join: keep only genes that are both DEGs and in tumor modules
overlap_genes <- inner_join(degs, module_genes, by = "gene_id_clean")

cat("Total DEGs:                   ", nrow(degs), "\n")
cat("Genes in tumor modules:       ", nrow(module_genes), "\n")
cat("Overlap (candidates for PPI): ", nrow(overlap_genes), "\n\n")

table(overlap_genes$module)

write.csv(overlap_genes,
          "07_functional_analysis/WGCNA_DEG_overlap.csv",
          row.names = FALSE)
```

# 11. Hub Gene Identification
```{r hub-genes}
MEs <- moduleEigengenes(datExpr_filtered, net$colors)$eigengenes

# Gene significance (GS): correlation of each gene with the Tumor trait
GS_tumor        <- as.numeric(cor(datExpr_filtered, traitData$Tumor, use = "p"))
names(GS_tumor) <- colnames(datExpr_filtered)

# Module membership (kME): correlation of each gene with its module eigengene
kME_blue      <- as.numeric(cor(datExpr_filtered, MEs$MEblue,      use = "p"))
kME_turquoise <- as.numeric(cor(datExpr_filtered, MEs$MEturquoise, use = "p"))
kME_brown     <- as.numeric(cor(datExpr_filtered, MEs$MEbrown,     use = "p"))

names(kME_blue) <- names(kME_turquoise) <- names(kME_brown) <- colnames(datExpr_filtered)

# Hub gene criteria: kME > 0.8 AND |GS| > 0.8
hub_blue      <- names(which(kME_blue      > 0.8 & abs(GS_tumor) > 0.8))
hub_turquoise <- names(which(kME_turquoise > 0.8 & abs(GS_tumor) > 0.8))
hub_brown     <- names(which(kME_brown     > 0.8 & abs(GS_tumor) > 0.8))

cat("Hub genes — blue:      ", length(hub_blue), "\n")
cat("Hub genes — turquoise: ", length(hub_turquoise), "\n")
cat("Hub genes — brown:     ", length(hub_brown), "\n")

all_hubs <- unique(c(hub_blue, hub_turquoise, hub_brown))
cat("Total unique hub genes:", length(all_hubs), "\n")
```

# 12. Convert Hub Gene IDs and Save for STRING
```{r convert-hub-ids}
library(org.Hs.eg.db)
library(AnnotationDbi)

# Strip version suffixes
all_hubs_clean <- sub("\\..*", "", all_hubs)

# Convert Ensembl to gene symbols
hub_symbols <- mapIds(
  org.Hs.eg.db,
  keys      = all_hubs_clean,
  column    = "SYMBOL",
  keytype   = "ENSEMBL",
  multiVals = "first"
)

hub_df <- data.frame(
  ensembl_id  = all_hubs_clean,
  gene_symbol = hub_symbols,
  module      = ifelse(all_hubs %in% hub_blue, "blue",
                ifelse(all_hubs %in% hub_turquoise, "turquoise", "brown")),
  row.names   = NULL
) %>%
  filter(!is.na(gene_symbol))

cat("Hub genes with symbol assigned:", nrow(hub_df), "\n")
head(hub_df, 20)

write.csv(hub_df,
          "07_functional_analysis/WGCNA_hub_genes.csv",
          row.names = FALSE)
```

# 13. STRING Input List
```{r string-list}
# One gene per line — STRING accepted format
string_input <- unique(hub_df$gene_symbol)

cat("Total genes for STRING:", length(string_input), "\n")

writeLines(string_input,
           "07_functional_analysis/hub_genes_for_STRING.txt")

cat("Saved to: 07_functional_analysis/hub_genes_for_STRING.txt\n\n")

# Breakdown by module
hub_df %>%
  group_by(module) %>%
  summarise(n_genes = n(),
            genes   = paste(gene_symbol, collapse = ", ")) %>%
  print()
```

# 14. PPI Consensus Hub Genes
```{r consensus-hubs}
library(readr)

# Load cytoHubba top-20 rankings from four algorithms
degree <- read_csv("07_functional_analysis/first20_DegreeModel.csv", skip = 1)
mcc    <- read_csv("07_functional_analysis/first20_MCCModel.csv",    skip = 1)
mnc    <- read_csv("07_functional_analysis/first20_MNCModel.csv",    skip = 1)
epc    <- read_csv("07_functional_analysis/first20_EPCModel.csv",    skip = 1)

genes_degree <- degree$Name
genes_mcc    <- mcc$Name
genes_mnc    <- mnc$Name
genes_epc    <- epc$Name

# Count how many algorithms rank each gene in the top 20
all_genes <- unique(c(genes_degree, genes_mcc, genes_mnc, genes_epc))

consensus_df <- data.frame(
  gene_symbol = all_genes,
  in_degree   = as.integer(all_genes %in% genes_degree),
  in_mcc      = as.integer(all_genes %in% genes_mcc),
  in_mnc      = as.integer(all_genes %in% genes_mnc),
  in_epc      = as.integer(all_genes %in% genes_epc)
) %>%
  mutate(n_algorithms = in_degree + in_mcc + in_mnc + in_epc) %>%
  filter(n_algorithms >= 3) %>%   # consensus threshold: top-20 in at least 3 algorithms
  arrange(desc(n_algorithms), gene_symbol)

cat("Consensus hub genes (>=3 algorithms):", nrow(consensus_df), "\n\n")
print(consensus_df)

write.csv(consensus_df,
          "07_functional_analysis/PPI_consensus_hub_genes.csv",
          row.names = FALSE)
```

# 15. Module Dendrogram
```{r dendrogram, fig.width=14, fig.height=6}
plotDendroAndColors(
  net$dendrograms[[1]],
  net$colors[net$blockGenes[[1]]],
  "Module colors",
  dendroLabels = FALSE,
  hang         = 0.03,
  addGuide     = TRUE,
  guideHang    = 0.05,
  main         = "Gene Dendrogram and Module Colors - Block 1"
)
```

# 16. Gene Significance vs Module Membership Scatter Plots
```{r gs-kme-plots, fig.width=12, fig.height=4}
# High kME + high GS = biologically relevant hub gene
par(mfrow = c(1, 3))

verboseScatterplot(
  kME_blue[net$colors == "blue"],
  GS_tumor[net$colors == "blue"],
  xlab   = "Module Membership (kME) - Blue",
  ylab   = "Gene Significance (GS) for Tumor",
  main   = "Blue Module",
  col    = "blue", abline = TRUE
)

verboseScatterplot(
  kME_turquoise[net$colors == "turquoise"],
  GS_tumor[net$colors == "turquoise"],
  xlab   = "Module Membership (kME) - Turquoise",
  ylab   = "Gene Significance (GS) for Tumor",
  main   = "Turquoise Module",
  col    = "turquoise", abline = TRUE
)

verboseScatterplot(
  kME_brown[net$colors == "brown"],
  GS_tumor[net$colors == "brown"],
  xlab   = "Module Membership (kME) - Brown",
  ylab   = "Gene Significance (GS) for Tumor",
  main   = "Brown Module",
  col    = "brown", abline = TRUE
)
```

# 17. Save Full Module Summary
```{r save-results}
module_summary <- data.frame(
  gene_id       = colnames(datExpr_filtered),
  module        = net$colors,
  kME_blue      = kME_blue,
  kME_turquoise = kME_turquoise,
  kME_brown     = kME_brown,
  GS_tumor      = GS_tumor
)

write.csv(module_summary,
          "07_functional_analysis/WGCNA_module_summary.csv",
          row.names = FALSE)

cat("=== WGCNA SUMMARY ===\n")
cat("Samples analyzed:         ", nrow(datExpr_filtered), "\n")
cat("Genes analyzed:           ", ncol(datExpr_filtered), "\n")
cat("Soft power selected:      ", softPower, "\n")
cat("Modules identified:       ", length(unique(net$colors)) - 1, "\n")
cat("Tumor modules selected:   blue, turquoise, brown\n")
cat("WGCNA hub genes:          ", length(all_hubs), "\n")
cat("Hub genes with symbol:    ", nrow(hub_df), "\n")
cat("Consensus hub genes (PPI):", nrow(consensus_df), "\n")
```

# 18. Save WGCNA Objects for Reuse
```{r save-wgcna}
# Save network object
saveRDS(net, "07_functional_analysis/WGCNA_net.rds")

# Save all objects needed by Script 08
save(sft, MEs, moduleTraitCor, textMatrix, GS_tumor,
     kME_blue, kME_turquoise, kME_brown,
     datExpr_filtered, traitData,
     file = "07_functional_analysis/WGCNA_objects.RData")

cat("WGCNA objects saved\n")
```

# 19. Hub Gene Bar Chart
```{r hub-barchart, fig.width=8, fig.height=6}
library(tidyverse)

degree <- read_csv("07_functional_analysis/first20_DegreeModel.csv", skip = 1)

# Consensus hub gene list (from Section 14)
hub_genes <- c("AIF1","CCR1","CD163","CD33","CD36","CD86","CTSS","CYBB",
               "TLR7","FCER1G","FCGR1A","FCGR3A","MYD88","TLR4","MNDA",
               "LILRB2","TLR8","C5AR1","FPR1")

# Use Degree scores as the ranking reference
degree_scores <- degree %>%
  filter(Name %in% hub_genes) %>%
  rename(gene = Name, score = Score) %>%
  arrange(desc(score))

p <- ggplot(degree_scores, aes(x = reorder(gene, score), y = score)) +
  geom_bar(stat = "identity", fill = "#4A90D9") +
  coord_flip() +
  labs(title = "Consensus Hub Genes — Degree Score",
       x     = NULL,
       y     = "Degree Score") +
  theme_classic(base_size = 12)

ggsave("08_figures_tables/article_figures/Fig4B_HubGenes_Degree.png",
       p, width = 8, height = 6, dpi = 300)

p
```
